\section{Symbolic and Neuro-Symbolic Monitoring Strategies}

\subsection{Symbolic Monitoring}
\label{sec:symbolic}

\paragraph{Architecture.}
The purely symbolic strategy is the classical automata-theoretic approach to runtime monitoring, and it serves as the exact baseline against which the two neuro-symbolic paradigms are measured. Given a specification $\varphi$, we compile it once into the minimal DFA $A_\varphi = (2^\Sigma, Q, q_0, \delta, F)$ introduced in Section~\ref{sec:ltlf}, using the standard \ltlf-to-automaton route \citep{LTLf,LTL2DFA1} as implemented by \texttt{ltlf2dfa} \citep{fuggitti-ltlf2dfa}. Monitoring is then a deterministic walk over this automaton: the monitor keeps a single current state $q$, initialized to $q_0$, and on observing the cell $\sigma_t \in 2^\Sigma$ it advances $q \leftarrow \delta(q, \sigma_t)$. Because the automaton is fixed and deterministic, each step costs a single transition lookup, independent of the trace seen so far and of the size of $\varphi$; after processing a prefix $\sigma_1 \cdots \sigma_t$ the current state is exactly $\delta^*(q_0, \sigma_1 \cdots \sigma_t)$. This makes symbolic monitoring the fastest paradigm on a single trace, at the cost of being entirely static: the automaton encodes the specification exactly and admits no learning or adaptation.

\paragraph{Verdicts and early termination.}
Online monitoring requires the three-valued runtime semantics of \citet{monitorability_LTL}: at each cell the monitor emits \textsc{satisfy}, \textsc{violate}, or \textsc{undecided}, the last meaning that the observed prefix can still be extended into both an accepting and a non-accepting continuation. A definite verdict can be issued as soon as the outcome is settled regardless of the future, which corresponds to two structural properties of the current state, both precomputed once by backward reachability over $A_\varphi$. A state is a \emph{trap} if no accepting state is reachable from it: reaching a trap means $\varphi$ is permanently violated, so the monitor halts with \textsc{violate}. Dually, a state is an \emph{accepting sink} if every state reachable from it (including itself) is accepting: the monitor halts with \textsc{satisfy}. With trap and sink sets materialized at compile time, the per-step verdict reduces to two set-membership tests on top of the transition lookup. As long as the current state is neither, the verdict is \textsc{undecided} and monitoring continues.

Some specifications never reach a trap or a sink, so \textsc{step} returns \textsc{undecided} for the entire trace; the canonical example is the response pattern $G(a \rightarrow Fb)$, which can always be satisfied by a later $b$ and always be rescued from violation, hence has neither an accepting sink nor a trap. For these the verdict is only binary at the end of the trace, and the monitor reports a final verdict by testing acceptance of the state reached, i.e.\ \textsc{satisfy} iff $\delta^*(q_0, \boldsymbol{\sigma}) \in F$ and \textsc{violate} otherwise. Crucially, this monitor is exact on every \ltlf formula: its verdicts coincide with the semantics of Section~\ref{sec:ltlf} by construction, including the nested-temporal specifications on which RuleRunner diverges (Section~\ref{sec:rulerunner}).

\subsection{Rule-based NeSy Monitoring: RuleRunner}
\label{sec:rulerunner}

\paragraph{Architecture.}

A RuleRunner system is a tuple $\langle S,R_E,R_R \rangle$, where $S$ is the state of the system, $R_E$ is the set of \emph{evaluation rules}, and $R_R$ is the set of \emph{reactivation rules}. In details:

\begin{itemize}
    \item The sate $S$ contains observations, active rules, and truth evaluations. $R[\phi]$ denotes an active rule, while $[\phi]V$  denotes a 3-values truth evaluation, with $V \in \{T, F, ?\}$.
    \item Evaluation rules are used to compute the truth value of a formula $\phi$ under verification, given the truth values of its direct subformulae $\psi_i$. They have the form $R[\phi],[\psi_1]V,...,[\psi_n]V \rightarrow [\phi]V$. For instance, $R[a \lor b], [a]T \rightarrow [a \lor b]T$, because the disjuction is evaluated to \emph{true} if one of the two disjuncts is true.
    \item Reactivation rules are used when a formula $\phi$ is evaluated to undecided, i.e. $[\phi]?$, in which case that formula and its subformulae are scheduled to be monitored again in the next cell of the trace. They have the form $[\phi]? \rightarrow R[\phi],R[\psi_1],...,R[\psi_n]$. For instance, $[\Diamond b]? \rightarrow R[b],R[\Diamond b]$, because if $\Diamond b$ is undecided, we need to re-evaluate $b$ and $\Diamond b$ at the next step.
\end{itemize}

%\noindent
%Note that evaluation rules are Horn clauses in implication form, and reactivation rules can be rewritten to be a set of Horn clauses as well.

\noindent
Both active rules and truth evaluations can be enriched by a \emph{qualifier}, such as $B$ (binary), $L$ (left) and $R$ (right), which provides additional information recording which operands within a formula/subformula are still being watched. For instance, $R[a \land \Diamond b], [a]T \rightarrow [a \land \Diamond b] ? R$, because if the left conjunct $a$ is true, it is still needed to check the truth value right conjunct $\Diamond b$ to assign a truth value to the conjunction.

%$\mathsf{B}$ (both operands), and $\mathsf{L}$/$\mathsf{R}$ (only the left/right operand, after the other has been settled to its pinning value); the next operator switches to a monitoring mode $\mathsf{M}$ that mirrors its operand one cell later. 

Verification of a trace against a formula proceeds iteratively on the cells of the trace. At the beginning of each iteration (corresponding to the monitoring of a cell), the state $S$ contains a set of rule names corresponding to the subformulae to be checked in that cell. First, the procedure includes observations of the current cell in the state $S$ and fires the evaluation rules repeatedly incrementing the state itself until it reaches a fixpoint. In practice, this means computing the truth values of all subformulae of $\phi$ in a bottom-up way, from simple atoms to $\phi$ itself (see Figure ...). If the root evaluates to true (resp. false), the monitor halts with verdict \textsc{satisfy} (resp.\ \textsc{violate}). Otherwise, the reactivation rules are fired, and the result constitute the state of the next iteration. Note that in this case the state is not expanded, but replaced by the result of reactivation rules. This is used to flush all the previous truth evaluation, which are to be computed from scratch in the new cell. The procedure terminates with a decided verdict because of special END-triggered evaluation rules, that returns either true or false if we consumed the whole trace. 

A RuleRunner system can be translated in a logic program, which in turn can be encoded in a Recurrent Neural Network with a single hidden layer, s.t. each evaluation and reactivation rule is a mapping from the input layer to the output one. For a detailed description of RuleRunner and verification algorithm see \citet{perotti2014neural,perotti2015runtime}.


\paragraph{Examples.}
A limitation of RuleRunner emerges for formulas in which a temporal operator is nested inside the operand of another temporal operator that may reactivate that operand across successive cells. RuleRunner relies on maintaining a single shared state, with one truth status for each syntactic subformula. This design is well suited to many flat temporal specifications, but in nested cases it can conflate distinct live obligations corresponding to the same subformula at different trace positions. 

Consider the two formulas $\varphi_g = F(a \land b)$ and $\varphi_b = F(a \land Xb)$, which differ only by a single next operator, monitored over the same trace $\boldsymbol{\sigma} = (\{a\}, \emptyset, \{b\})$. Neither formula is satisfied: $\varphi_g$ would require a cell where $a$ and $b$ hold \emph{together} (but $a$ occurs only at $t = 0$ and $b$ only at $t = 2$), and $\varphi_b$ would require a cell $t$ with $a \in \sigma_t$ and $b \in \sigma_{t+1}$ (but the only $a$ is at $t = 0$,and is not followed by a $b$ at $t = 1$). The correct verdict in both cases is therefore \textsc{violate}, but RuleRunner can provide the correct answer only for the first formula.  

First, consider $\varphi_g$. The run is:
%
\noindent
\begin{center}\small
\begin{tabular}{c c l l}
\toprule
$t$ & $\sigma_t$ & evaluation & reactivation\\
\midrule
\multirow{2}{*}{$0$} & \multirow{2}{*}{$\{a\}$} & $[a]T,[b]F \rightarrow [a \land b]F$ & \\
& & $[a \land b]F \rightarrow [F(a \land b)]?$ & $\rightarrow R[F(a \land b)],R[a],R[b],R[a \land b]$\\[1.5ex]
\multirow{2}{*}{$1$} & \multirow{2}{*}{$\emptyset$} & $[a]F,[b]F \rightarrow [a \land b]F$ & \\
& & $[a \land b]F \rightarrow [F(a \land b)]?$ & $\rightarrow R[F(a \land b)],R[a],R[b],R[a \land b]$\\[1.5ex]
\multirow{3}{*}{$2$} & \multirow{3}{*}{$\{b\}$} & $[a]F,[b]T \rightarrow [a \land b]F$ & \\
& & $[a \land b]F \rightarrow [F(a \land b)]?$ & \\
& & $[F(a \land b)]? \rightarrow_{\textsc{end}} [F(a \land b)]F$ & $\rightarrow \textsc{violate}$ \\
\bottomrule
\end{tabular}
\end{center}

Now, consider $\varphi_b$, In this case, the run diverges:
\begin{center}\small
\begin{tabular}{c c l l}
\toprule
$t$ & $\sigma_t$ & evaluation & reactivation\\
\midrule
\multirow{3}{*}{$0$} & \multirow{3}{*}{$\{a\}$} & $R[Xb] \rightarrow [Xb]?$ & $\rightarrow R[Xb]M,R[b]$ \\
&  & $[a]T,[Xb]? \rightarrow [a \land Xb]?R$ & $\rightarrow R[a \land Xb]R$  \\
&  & $[a \land Xb]?R \rightarrow [F(a \land Xb)]?$ & $\rightarrow R[F(a \land Xb)],R[a],R[Xb],R[a \land Xb]$ \\[1.5ex]

\multirow{6}{*}{$1$} & \multirow{6}{*}{$\emptyset$} & $R[Xb] \rightarrow [Xb]?$ & $\rightarrow R[Xb]M,R[b]$ \\
& & $R[Xb]M, [b]F \rightarrow [Xb]F$ & \\
&  & $R[a \land Xb]R,[Xb]? \rightarrow [a \land Xb]?R$ & $\rightarrow R[a \land Xb]R$  \\
&  & $R[a \land Xb]R,[Xb]T \rightarrow [a \land Xb]T$ &  \\
&  & $R[a \land Xb],[a]F \rightarrow [a \land Xb]F$ &  \\
&  & $[a \land Xb]?R \rightarrow [F(a \land Xb)]?$ & $\rightarrow R[F(a \land Xb)],R[a],R[Xb],R[a \land Xb]$ \\[1.5ex]

\multirow{5}{*}{$2$} & \multirow{5}{*}{$\{b\}$} & $R[Xb] \rightarrow [Xb]?$ & \\
& & $R[Xb]M, [b]T \rightarrow [Xb]T$ & \\
&  & $R[a \land Xb]R,[Xb]T \rightarrow [a \land Xb]T$ &  \\
&  & $R[a \land Xb],[a]F \rightarrow [a \land Xb]F$ &  \\
&  & $[a \land Xb]T \rightarrow [F(a \land Xb)]T$ & $\rightarrow \textsc{satisfy}$ (wrong) \\
\bottomrule
\end{tabular}
\end{center}

\noindent
The divergence stems from RuleRunner's single shared state, which keeps exactly one truth slot per syntactic subformula. For $\varphi_g$ the operand $a \land b$ is purely propositional: it is re-evaluated from scratch at every cell and never carried across cells, so the conjunction always reflects the current instant alone, and the monitor correctly reports \textsc{violate}. For $\varphi_b$, instead, the operand contains the temporal subformula $Xb$, whose truth is only settled one cell later; the eventually operator spawns a fresh copy of $a \land Xb$ at every cell, yet all copies share the single literal $[Xb]$ and the single carried conjunction $[a \land Xb]R$. When $b$ finally holds at $t = 2$, the resulting $[Xb]T$ is consumed by the carried mode-$R$ conjunction, which pins its already-settled left conjunct to true and concludes $[a \land Xb]T$ --- even though no single instant ever had $a$ immediately followed by $b$. Lacking any way to distinguish obligations originating at different positions, RuleRunner conflates them and wrongly reports \textsc{satisfy}. This is a structural consequence of the one-literal-per-subformula encoding, not an implementation artifact, and it affects exactly the nested-temporal formulas (such as the response pattern $G(a \rightarrow Fb)$) that are common in practice; the automata-based paradigms of the next section do not suffer from it, since a single canonical state machine never conflates concurrent obligations.



\subsection{Automata-based NeSy Monitoring: DeepDFA}
\label{sec:deepdfa}

\paragraph{Architecture.}
DeepDFA \citep{deepdfa_ecai2024} keeps the automaton of the symbolic paradigm but replaces its explicit state walk with a differentiable, tensorized one; we adopt the same forward formulation used for predictive monitoring of business processes by \citet{mezini2025neuro}. Starting again from the minimal DFA $A_\varphi = (2^\Sigma, Q, q_0, \delta, F)$ of Section~\ref{sec:ltlf}, we encode its transition function as a three-way tensor $T \in \{0,1\}^{|Q| \times |2^\Sigma| \times |Q|}$ with $T[q, \sigma, q'] = 1$ iff $\delta(q, \sigma) = q'$, and the accepting set as an indicator vector $v_o \in \{0,1\}^{|Q|}$ with $v_o[q] = 1$ iff $q \in F$. The monitor state is no longer a single automaton state but a row vector $\mathbf{q}_t \in [0,1]^{|Q|}$ over states, initialized to the one-hot vector $\mathbf{q}_0 = v_i$ of $q_0$. Writing $T_\sigma = T[\,\cdot\,, \sigma, \cdot\,] \in \{0,1\}^{|Q| \times |Q|}$ for the (row-stochastic) transition matrix of symbol $\sigma$, a step is a single vector--matrix product
\begin{equation}
  \mathbf{q}_t = \mathbf{q}_{t-1}\, T_{\sigma_t},
  \qquad
  \mathrm{acc}_t = \mathbf{q}_t \cdot v_o,
  \label{eq:deepdfa-crisp}
\end{equation}
where $\mathrm{acc}_t \in [0,1]$ is the acceptance score of the prefix read so far. On crisp observations every $T_{\sigma_t}$ is a $0/1$ permutation-like matrix and $\mathbf{q}_t$ stays one-hot, so the update reproduces the symbolic walk exactly: $\mathbf{q}_t$ is the indicator of $\delta^*(q_0, \sigma_1 \cdots \sigma_t)$. Consequently DeepDFA is, like the symbolic monitor and unlike RuleRunner, \emph{exact on every} \ltlf \emph{formula}, including the nested-temporal specifications of Section~\ref{sec:rulerunner}; the single canonical automaton again precludes the conflation of concurrent obligations. The three-valued verdict and early termination are read off exactly as in Section~\ref{sec:symbolic}, by testing membership of $\argmax_q \mathbf{q}_t[q]$ in the precomputed trap and accepting-sink sets, and the final verdict tests its acceptance.

\paragraph{Soft observations and differentiability.}
The reason to pay for this tensorization is that Equation~\eqref{eq:deepdfa-crisp} extends verbatim to \emph{probabilistic} observations, which the symbolic walk cannot represent. Suppose a cell is delivered not as a crisp assignment but as a vector of per-atom probabilities $\mathbf{p}_t \in [0,1]^{\Sigma}$, as produced by a neural perceptor or a noisy sensor. We replace the one-hot symbol selection by an \emph{expected} transition matrix $M(\mathbf{p}_t) \in [0,1]^{|Q| \times |Q|}$, whose entry $(q, q')$ is the probability that the guard of edge $q \to q'$ is satisfied under $\mathbf{p}_t$, computed compositionally over the boolean structure of the guard under an atom-independence assumption ($P(a) = p_a$, $P(\lnot \psi) = 1 - P(\psi)$, $P(\psi_1 \land \psi_2) = P(\psi_1)P(\psi_2)$, $P(\psi_1 \lor \psi_2) = 1 - (1 - P(\psi_1))(1 - P(\psi_2))$). The update and acceptance score are unchanged,
\begin{equation}
  \mathbf{q}_t = \mathbf{q}_{t-1}\, M(\mathbf{p}_t),
  \qquad
  \mathrm{acc}_t = \mathbf{q}_t \cdot v_o,
  \label{eq:deepdfa-soft}
\end{equation}
but now $\mathbf{q}_t$ is a genuine distribution over DFA states and $\mathrm{acc}_t$ is a calibrated acceptance probability that propagates the perceptual uncertainty all the way to the verdict, rather than discarding it through a hard $0.5$ threshold. Because each $M(\mathbf{p}_t)$ is a smooth, row-stochastic function of $\mathbf{p}_t$ (and, if desired, of learnable parameters of $T$), the whole monitor is differentiable: gradients of a verdict loss flow back through the automaton, which is what makes DeepDFA the vehicle for gradient-based specification adaptation and, ultimately, end-to-end training of a perceptor. On crisp inputs $M(\mathbf{p}_t)$ collapses to $T_{\sigma_t}$ and the probabilistic semantics is exact for any guard; on fractional inputs it is exact whenever the guard is \emph{read-once} (each atom occurring once), and an independence approximation otherwise.

\paragraph{The alphabet blow-up.}
DeepDFA's structural cost is dual to those of the other two paradigms. The transition tensor is indexed by the full alphabet $2^\Sigma$, so a dense materialization is exponential in the number of atoms, $|T| = |Q|^2 \cdot 2^{|\Sigma|}$. One can avoid building the dense tensor by evaluating the per-edge guard probabilities on the fly --- the factored form underlying Equations~\eqref{eq:deepdfa-crisp}--\eqref{eq:deepdfa-soft}, which never enumerates $2^\Sigma$ --- and this is what we use for specifications over many atoms. But the blow-up is not merely an implementation choice: when a single guard depends on all $|\Sigma|$ atoms there is in general no sub-exponential set of symbols to sum over, and only the read-once circuit structure of a guard (which a flat DFA does not expose) permits its satisfaction probability to be computed in time linear in $|\Sigma|$. This alphabet blow-up is therefore DeepDFA's Achilles heel, the counterpart of the symbolic monitor's state-space blow-up and RuleRunner's nested-temporal limit, and we return to it in the comparison of Section~\ref{sec:theory-cmp} and the scalability experiments of Section~\ref{sec:experiments}.
